Regression-based

### 2023-12-15 Final Project.Rmd
refined the models

# ASK MIHIR TO PUT IN IMAGES HERE !!!

## Analysis Objectives

@@

A link to the original study is here: https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf

### Load in Packages
```{r}

@@

### Data Processing
```{r}
#load in the data
library(tidyverse)
#load data
cirrhosis <- read_csv("cirrhosis.csv")
head(cirrhosis)
```

### Pre-process the data
```{r}
#encode the sex, ascites, hepatomegaly, spiders, edema into binary variables
cirrhosis$Sex <- ifelse(cirrhosis$Sex == "F", 0, 1)
cirrhosis$Ascites <- ifelse(cirrhosis$Ascites == "Y", 1, 0)
cirrhosis$Hepatomegaly <- ifelse(cirrhosis$Hepatomegaly == "Y", 1, 0)
cirrhosis$Spiders <- ifelse(cirrhosis$Spiders == "Y", 1, 0)
#factorize the stage
cirrhosis$Stage <- factor(cirrhosis$Stage)
cirrhosis$Status <- factor(cirrhosis$Status)
cirrhosis$Drug <- factor(cirrhosis$Drug)
cirrhosis$Sex <- factor(cirrhosis$Sex)
#cirrhosis$Ascites <- factor(cirrhosis$Ascites)
#cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
#cirrhosis$Spiders <- factor(cirrhosis$Spiders)
cirrhosis$Edema <- factor(cirrhosis$Edema)
cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))

@@

- Use Machine Learning Algorithms:
  - You must use at least one of each: a Regression-based algorithm, a Tree-based algorithm, and an Unsupervised Learning algorithm.
  - Each algorithm must be explained for a general audience.
  - All results should be explained.
  - Where appropriate, comparisons between methods should be discussed. This includes highlighting the strengths and weaknesses of the various approaches as well as when each is appropriate.

- A discussion of what you’ve learned from your work about the context and recommendations for future work.

- Author Contributions–list out who did what in the project.

- References–Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.

>>>>

![[Primary Biliary Cholangitis ([Source](https://www.google.com/search?client=safari&sca_esv=591276386&rls=en&sxsrf=AM9HkKmtJVOC2Jrf2BSYVB2O-QcnTSgx6g:1702664705100&q=Primary+Biliary+Cholangitis&tbm=isch&source=lnms&sa=X&ved=2ahUKEwid0qyniJKDAxX9FFkFHVH3CcoQ0pQJegQIDxAB&biw=1440&bih=848&dpr=2#imgrc=UeV_fhP-8vPOlM))]{.underline}](17715-primary-biliary-cholangitis.webp)

## Objectives

@@

A link to the original study is here: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>

### Load in Packages

@@

### Data Processing

```{r}
#load in the data
library(tidyverse)
#load data
cirrhosis <- read_csv("cirrhosis.csv")
head(cirrhosis)
```

### Pre-process the data

```{r}
#encode the sex, ascites, hepatomegaly, spiders, edema into binary variables
cirrhosis$Sex <- ifelse(cirrhosis$Sex == "F", 0, 1)
cirrhosis$Ascites <- ifelse(cirrhosis$Ascites == "Y", 1, 0)
cirrhosis$Hepatomegaly <- ifelse(cirrhosis$Hepatomegaly == "Y", 1, 0)
cirrhosis$Spiders <- ifelse(cirrhosis$Spiders == "Y", 1, 0)
#factorize the stage
cirrhosis$Stage <- factor(cirrhosis$Stage)
cirrhosis$Status <- factor(cirrhosis$Status)
cirrhosis$Drug <- factor(cirrhosis$Drug)
cirrhosis$Sex <- factor(cirrhosis$Sex)
#cirrhosis$Ascites <- factor(cirrhosis$Ascites)
#cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
#cirrhosis$Spiders <- factor(cirrhosis$Spiders)
cirrhosis$Edema <- factor(cirrhosis$Edema)
cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))

@@

-   Use Machine Learning Algorithms:

    -   You must use at least one of each: a Regression-based algorithm, a Tree-based algorithm, and an Unsupervised Learning algorithm.
    -   Each algorithm must be explained for a general audience.
    -   All results should be explained.
    -   Where appropriate, comparisons between methods should be discussed. This includes highlighting the strengths and weaknesses of the various approaches as well as when each is appropriate.

-   A discussion of what you've learned from your work about the context and recommendations for future work.

-   Author Contributions--list out who did what in the project.

-   References--Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.

### 2023-12-15 Regression.Rmd
refined the models

# Part II: Exploring the predictors for a Liver Transplant using logistic regression

Next, let us explore our question with a different type of model. To do this, we will be using (binary) logistic regression model

@@

![Logistic Regression Model ([Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/))](46-4.png)

@@

@@

### Logistic regression model

@@

# Set seed for reproducibility and slice
set.seed(380)
trainingData <- LRmodelData %>%
  group_by(Status) %>%  # group_by() function ensures that the data
  slice_sample(prop = 0.75)

@@

```{r}
# Form Candidate Model 1
model1 <- glm(
  formula = Status ~ Platelets,
  data = trainingData,
  family = binomial
)
```

@@

Initially, we'll delve into the two preliminary models independently to
understand where they stand. Following that, we'll be deploying the best
candidate model on our test data. Regarding confusion matrices, we'll
employ a basic rule: if the predicted probability of a student
graduating exceeds 0.5, we'll categorize them as having graduated (naïve
rule).

@@

```{r}
# Model 1 Coefficient Table
as.data.frame(summary(model1)$coefficients) %>%
  rownames_to_column(var = "term") %>%
  rename(coefficient = Estimate) %>% 
  mutate(
    prob_odds = case_when(
      coefficient == "(Intercept)" ~ exp(coefficient)/(1 + exp(coefficient)),
      .default = exp(coefficient)
    ),
    .after = coefficient
  ) %>%
  mutate(
    `Pr(>|z|)` = ifelse(
      test = `Pr(>|z|)` < 0.001,
      yes = paste("< 0.001"),
      no = `Pr(>|z|)`
    ),
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      grepl(x = term, pattern = "Platelets") ~ "Platelets"
    )
  ) %>%
  kable()

@@

```{r}
library(janitor)
# Building confidence intervals for Model 1 coefficients
model1CI <- confint(
  object = model1,
  parm = "Platelets",
  level = 0.9
)

@@

# Apply naïve rule ----
trainingResults <- trainingResults %>%
  mutate(
    model1Class = case_when(
      model1Pred > 0.5 ~ 1,
      .default = 0
    )
  )

#Confusion Matrix for Model 1
trainingResults %>%
  mutate(Patient_status = ifelse(Status == 1, 1, 0)) %>%
  tabyl(var1 = model1Class, var2 = Patient_status) %>%
  adorn_title(
    placement = "combined",
    row_name = "Predicted",
    col_name = "Actual"
  ) %>%
  kable(
    booktabs = TRUE,
    align = "c",
    caption = "Model 1 Confusion Matrix"
  )

@@

Next let us look at our Model 2, which has multiple factors as discussed
earlier.

@@

@@

@@

Here we can see the generalized variance inflation factors for candidate
Model 2. Multicollinearity is a statistical concept where several
independent variables in a model are correlated and this is an important
consideration especially when we deal with so many categories.

```{r}
#Store the predicted and actual values for Model 2
trainingResults$model2Pred <- predict(model2, type = "response")
trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, 1, 0)
trainingResults$Actual <- ifelse(trainingData$Status == 1, 1, 0)

# Create confusion matrix using table
confusionMatrix <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)

kable(confusionMatrix, caption = "Confusion matrix for Model 2") %>%
  kable_classic(latex_options = "HOLD_position")

@@

@@

```{r}
# Convert 'Actual' column to numeric 0/1
trainingResults <- trainingResults %>%
  mutate(
    actualNum = if_else(Actual == 0, 0, 1)
  )

@@

The separation plot assesses the the fit of the model by providing the
model's ability to predict occurrences with a high probability and
non-occurrences with low probability. The separation plot above
separation plot suggests that Model 2 has a reasonably good performance
in predicting graduates, especially for the observations on the
left-most side of the plot compared to Model 1 with the training data.
We will later compare this graph with the testing data.

@@

Model 1: - Its ROC curve is above the line of no discrimination,
indicating that the model has some predictive capabilities. - The AUC is
0.621, which is better than random guessing but suggests there's room
for improvement.

Model 2: - The ROC curve for Model 2 is significantly above that of
Model 1, and much closer to the top-left corner, indicating better
predictive performance. - The AUC is 0.768, which suggests a "good"
classification performance, and it's notably better than Model 1. - Its
ability to discriminate between positive and negative classes is
superior to that of Model 1.

@@

The influence plot shows several data points with high leverage and
large residuals, indicating potential outliers. Some observations,
notably those labeled like "23881," have significant influence on the
regression model due to their Cook's D values.

@@

```{r}
# Set up testing data results
testingData <- testingData %>%
  mutate(
    gradNum = case_when(
      Status == 0 ~ 0,
      Status == 1 ~ 1
    ),
    .after = Status
  )
testingData$predict <- predict(
  object = model2,
  newdata = testingData,
  type = "response"
)
testingData <- testingData %>%
  mutate(
    model2Class = case_when(
      predict > 0.5 ~ 1,
      .default = 0
    )
  )
```

```{r}
testingData$Status <- ifelse(testingData$Status == 1, 1, 0)

@@

From the matrix, we can see that the model has the following metrics:

-   True Positive (TP): Predicted Graduated and actually Graduated. In
    this case, 7251.
-   False Positive (FP): Predicted Graduated but actually Not Graduated.
    In this case, 3505.
-   False Negative (FN): Predicted Not Graduated but actually Graduated.
    In this case, 1936.
-   True Negative (TN): Predicted Not Graduated and actually Not
    Graduated. In this case, 5395.

From this, we can interpret that.

-   Accuracy: Approximately 69.92%
-   Precision: Approximately 67.41%
-   Recall (Sensitivity): Approximately 78.93%
-   F1 Score: Approximately 72.72%

This shows that our model is still capable with new data, with a slight
reduction to the overall performance.

>>>>

# Part II: Exploring the predictors for the patient's status using logistic regression

Next, let us explore our question with a different type of model and compare the results with our tree. To do this, we will be using (binary) logistic regression model.

@@

![[Logistic Regression Model ([Source](https://www.spiceworks.com/tech/artificial-intelligence/articles/what-is-logistic-regression/))]{.underline}](46-4.png)

@@

With this, we will build two candidate models:

-   The first model will test the classification based on just the SGOT

-   The second model will use a step wise function using various predictors to see the best performance.

Another important consideration is the application of prediction (estimating an outcome based on input variables) and inference (understanding the relationships between variables). We will evaluate the inference through the coefficient analysis and prediction through roc curves and confusion matrix. It is important to note that these metrics will complement each other in our understanding of the data. However, the main focus of this analysis will be prediction and we will work with several metrics to evaluate it.

We will use a variety of tools to understand the model's performance.

@@

@@

# Set seed for reproducibility and slice
set.seed(380)
trainingData <- LRmodelData %>%
  group_by(Status) %>%  # group_by() function ensures that the data
  slice_sample(prop = 0.80)

@@

```{r}
# Form Candidate Model 1
model1 <- glm(
  formula = Status ~ SGOT,
  data = trainingData,
  family = binomial
)
```

Stepwise results:

@@

Initially, we'll delve into the two preliminary models independently to understand where they stand. Following that, we'll be deploying the best candidate model on our test data. Regarding confusion matrices, we'll employ a basic rule: if the predicted probability of a the patient's status being "deceased" exceeds 0.5, we'll categorize them as deceased (naïve rule).

@@

```{r}
# Model 1 Coefficient Table
as.data.frame(summary(model1)$coefficients) %>%
  rownames_to_column(var = "X") %>%
  rename(coefficient = Estimate) %>% 
  mutate(
    prob_odds = case_when(
      coefficient == "(Intercept)" ~ exp(coefficient)/(1 + exp(coefficient)),
      .default = exp(coefficient)
    ),
    .after = coefficient
  ) %>%
  mutate(
    `Pr(>|z|)` = ifelse(
      test = `Pr(>|z|)` < 0.001,
      yes = paste("< 0.001"),
      no = `Pr(>|z|)`
    ),
    X = case_when(
      X == "(Intercept)" ~ "Intercept",
      grepl(x = X, pattern = "SGOT") ~ "SGOT"
    )
  ) %>%
  kable()

@@

This table shows us the results of our first model. We can see that, holding other variables constant, a one-unit increase in SGOT is associated with a decrease in the log-odds of the response variable by 0.0076446. Furthermore, the odds-ratio indicates that for each one-unit increase in SGOT, the odds of the event occurring decrease by about 0.76%.

We can also plot the confusion matrix for this model:

```{r}
library(janitor)
# Building confidence intervals for Model 1 coefficients
model1CI <- confint(
  object = model1,
  parm = "SGOT",
  level = 0.9
)

@@

# Apply naïve rule ----
trainingResults <- trainingResults %>%
  mutate(
    model1Class = case_when(
      model1Pred > 0.5 ~ "Censored",
      .default = "Deceased"
    )
  )

#Confusion Matrix for Model 1
trainingResults %>%
  mutate(Patient_status = ifelse(Status == 1, "Censored", "Deceased")) %>%
  tabyl(var1 = model1Class, var2 = Patient_status) %>%
  adorn_title(
    placement = "combined",
    row_name = "Predicted",
    col_name = "Actual"
  ) %>%
  kable(
    booktabs = TRUE,
    align = "c",
    caption = "Model 1 Confusion Matrix"
  )%>%kable_styling(latex_options = "HOLD_position")

@@

We can see that this model tends to over predict censored values. This shows the need of bringing in more factors. Next let us look at our Model 2, which has multiple factors as discussed earlier.

@@

This analysis suggests significant associations between several predictors (Bilirubin, Age, Alk_Phos, Prothrombin, Spiders) and the status of the patient. Each predictor, including the Intercept, shows a statistically significant effect, as indicated by their small p-values (Pr(\>\|z\|)). The negative coefficients for Bilirubin, Age, Alk_Phos, Prothrombin, and Spiders imply that increases in these variables are associated with a decrease in the log-odds of the outcome. Bilirubin and Spiders are showing notably strong effects in the model.

@@

This figure shows us the Tukey-Anscombe plot using Pearson residuals for Model 2. In an ideal fit, the residuals should be evenly distributed about zero with constant mean and variance. The shape of the line suggests that the model is not capturing someunder lying structure in the data, especially in extreme cases.

@@

The Variance Inflation Factor (VIF) values for the variables in the model (Bilirubin, Age, Alk_Phos, Prothrombin, Spiders) are all close to 1, indicating minimal multicollinearity. This means that these predictors are relatively independent of each other, enhancing the reliability of the model.

```{r}
#Store the predicted and actual values for Model 2
trainingResults$model2Pred <- predict(model2, type = "response")
trainingResults$model2Class <- ifelse(trainingResults$model2Pred > 0.5, "Censored", "Deceased")
trainingResults$Actual <- ifelse(trainingData$Status == 1, "Censored", "Deceased")

# Create confusion matrix using table
confusionMatrixRegression <- table(Predicted = trainingResults$model2Class, Actual = trainingResults$Actual)

kable(confusionMatrixRegression, caption = "Confusion matrix for Model 2") %>%
  kable_classic(latex_options = "HOLD_position")

@@

From this confusion matrix we can see the relationships between the True Positive, True Negative, False Positive and False Negative values. From this we can calculate:

-   **Accuracy**: Approximately 81.36%

-   **Recall**: Approximately 79.01%

-   **Precision**: Approximately 72.73%

-   **F1 Score**: Approximately 75.74%

@@

```{r}
# Convert 'Actual' column to numeric 0/1
trainingResults <- trainingResults %>%
  mutate(
    actualNum = if_else(Actual == "Deceased", 0, 1)
  )

@@

The separation plot assesses the the fit of the model by providing the model's ability to predict occurrences with a high probability and non-occurrences with low probability. The separation plot above separation plot suggests that Model 2 has a reasonably good performance in predicting the patient's status, especially for the observations on the left-most side of the plot compared to Model 1 with the training data. We will later compare this graph with the testing data.

@@

Model 1: - Its ROC curve is above the line of no discrimination, indicating that the model has some predictive capabilities. - The AUC is 0.64, which is better than random guessing but suggests there's room for improvement.

Model 2: - The ROC curve for Model 2 is significantly above that of Model 1, and much closer to the top-left corner, indicating better predictive performance. - The AUC is 0.899, which suggests a good classification performance, and it's notably better than Model 1. - Its ability to discriminate between positive and negative classes is superior to that of Model 1.

@@

The influence plot shows several data points with high leverage and large residuals, indicating potential outliers. Some observations, notably those labeled like "149," have significant influence on the regression model due to their Cook's D values.

@@

```{r}
# Set up testing data results
testingData <- testingData %>%
  mutate(
    gradNum = case_when(
      Status == 0 ~ 0,
      Status == 1 ~ 1
    ),
    .after = Status
  )
testingData$predict <- predict(
  object = model2,
  newdata = testingData,
  type = "response"
)
testingData <- testingData %>%
  mutate(
    model2Class = case_when(
      predict > 0.5 ~ "Censored",
      .default = "Deceased"
    )
  )
```

```{r}
testingData$Status <- ifelse(testingData$Status == 1, "Censored", "Deceased")

@@

-   **Accuracy**: Approximately 71.43%

-   **Recall**: Approximately 76.92%

-   **Precision**: Approximately 43.48%

-   **F1 Score**: Approximately 55.56%

This shows that our model is has struggled with overfitting, with an especially low precision score. We will discuss this in the comparison with the tree model, but this is a limitation of our data size. Lastly, we can plot the separation plot.
